### 2011-08-10 Paper/Paper.tex
Report finished.

% Multiline tables
\usepackage{array}
\usepackage{booktabs}

@@

% [DO NOT CHANGE] The addess of the educational programme.
University of Amsterdam\\
Faculty of Science\\
Science Park 904\\
1098 XH Amsterdam

@@

% [CHANGE] The address of the institute at which your supervisor is working.
% Be sure to include (1) institute (is appropriate), (2) faculty (if
% appropriate), (3) organisation name, (4) organisation address (2 lines).
Institute for Language and Logic\\
Faculty of Science\\
University of Amsterdam\\
Science Park 904\\
1098 XH  Amsterdam

@@

%Musical structure is extracted automatically. A dataset containing performances aligned to scores is then used to learn how structure relates to expression. A hidden markov model is used to model this relation. Finally, the system can be applied to a a new score to produce an expressive performance.

@@

{\bf Keywords:} performance rendering, musical structure, constituent structure, polyphonic piano music
\end{abstract}
\section{Introduction}

@@

Given a hypothetical perfect synthesizer, performing scores with computers is a trivial task. However \textit{expressively} performing music is not and much research has focussed on this issue. A wide selection of scientific fields may have interest in such research: Artificial intelligence, cognitive science and musicology all benefit from a better understanding of expression in music.

@@

A rather practical approach to expressive music performance is to recreate it. A performance rendering system is a computer program that automatically creates an expressive, humanlike performance of a musical score. Performance rendering systems have always faced a problem of evaluation. It is debatable how much we should value correlation of generated performances with human performances of the same piece. A Chopin interpretation of Arthur Rubinstein can be radically different from a Chopin interpretation of Vladimir Horowitz, yet both are considered high quality interpretations. This problem has been one of the motivations for a yearly performance rendering event where performance rendering systems compete and are judged by an audience.

Over the years, many performance rendering systems have been proposed. Early attempts, like Director Musices \cite{friberg2000generating}, have applied sets of rules to the problem. As the availablity of corpora grew, machine learning systems have been proposed as well. Gerhard Widmer is working with a large dataset containing almost all works by Chopin, performed by one pianist. A performance rendering system using this dataset called YQX \cite{yqx} was proposed by Sebastian Flossman et al. in 2008. This system won in al three categories of Rencon2008. At Rencon2011 it was ranked best in a blind evaluation by experts and ranked second in the evaluation by an audience. It should be noted that despite its age, Director Musices ranked third in the blind evaluation, suggesting that although performance rendering systems have advanced, the field is still in its infancy.

Many of the performance rendering systems that use machine learning, including YQX, predict expression at note level. Some do use structure. Pop-E\cite{hashida2006pop}, a rule based performance rendering system, uses a structure analysis based on Lerdahls and Jackendoff's grouping preference rules, but the structural analysis must be supplied to it by a human expert. In this thesis we will motivate the need for structure level expression and propose a system that purely uses structure level expression. The YQX system is very successful at the moment, therefore we will use this system as a reference for introducing our own system. In the following section we will provide a short analysis of some shortcomings of the YQX system and argue what we believe is the reason for these shortcomings.

What we mean by structure level expression and why we think it is important is clarified in section \ref{sec:structure}. Section \ref{sec:approach} will describe the system as a whole. Section \ref{sec:method} will describe every component of the system in detail. Practical issues are discussed in section \ref{sec:implementation}. Results are presented in section \ref{sec:results} and discussed in section \ref{sec:discussion}. Suggestions for improving the system and future research can be found in section \ref{sec:futurework}.

%Early performance rendering systems used rule-based approaches to recognise small local structures that are used as indications to add certain kind of expression. A prominent ongoing rule based system, \textit{Director Musices} uses 30 rules with weighting parameters that take music features as input and produce expressive actions as output. The rules where created with the help of musical experts. \textit{Pop-E} is a more recent rule based system that uses aspects of Lerdahl and Jackendoff's Generative Theory of Tonal Music(GTTM) [Introduce this]. Gerhard Widmer is doing research using a large dataset of performances aligned to scores.

@@

The YQX system, as Flossman et al. admit, tended to sometimes produce nervous sounding changes in expression. They present two extensions tailored towards generating smoother performances. The problem with these extensions is, as Flossman et al. also admit himself, is that the increased smoothness comes at the expense of expressitivity. To compensate for this, three explicit rules are added to postprocess the performances. We think the reason that Flossman et al. stumbled upon this tradeoff between expressiveness and nervousness is that the nervousness is inherent to note level performance rendering. Note level systems simply are not suited to capture one component of expression, namely structure level expression.

The system presented here will use structure exclusively to generate performances and will completely ignore note level expression. We hope that this system is able capture the sudden changes of expression that a note level system cannot easily learn. Sudden changes in expression occur frequently in expressive performances, a few notes may be played very loud, or very slow followed by soft notes played very fast. When done at the right moment, such sudden changes can greatly improve the humanlike feel of the performance, however when done at the wrong moment, they can ruin a performance. We think that at structure level it will be easier to capture the context in which sudden changes occur.

@@

\subsection{Structure Level Expression}
When listening to music, the listener's musical intuition assigns a certain hierarchical structure to the music: Notes make up phrases, phrases make up themes and themes make up a piece. In a performance, this structure may be accentuated in different ways. Accentuation happens at different levels, at note level performers may slow down at the end a phrase or introduce small pauses in the performance at phrase transitions. At constituent level one phrase may be played very loud, fast or staccato, while the next may be played slow, soft and legato.

To formally describe musical structure, we can look at music in a way similar to the way we look at natural language processing(NLP). In this analogy we see a piece of music as a sentence, which consists of constituents that individually can be made of constituents as well. We can recursively subdivide constituents in more constituents until we reach a terminal symbol. In NLP this may be a word, in music, this may be a note. We could represent musical structure as a parse tree. This paradigm corresponds to the intuition that a melody is not simply a sequence of notes but that notes form phrases. A phrase is always a constituent but a constituent is not always a phrase, it can be a distinguished part of a phrase or a distinguished set of phrases as well.

We must note that musical scores can be highly ambiguous and even experienced listeners may not agree on the correct parsetree of a piece. Quite often there may simply be more than one parse tree that makes musical sense. This should not be a problem for a performance rendering system: different expressive interpretations of one piece can be very diverse and still be accepted as sensible interpretations of the piece. As long as the parse tree does make at least some musical sense, a performance rendering system should be able to use it.

Although the YQX does have some notion of structure\footnote{One of the note features is distance to nearest point of closure}, expression is only predicted per note. The authors admit that the first simple version of the system ``tended to sometimes produce unstable, `nervous' sounding performances''. The only way to overcome this problem was to introduce methods that limited the expressiveness of performances. We consider this trade-off to be inherent to note-level expression based systems. To solve it, some notion of structure level expression is required.

@@

In this thesis, we propose a structure based performance rendering (SBPR) system. The system presented here ignores note level expression. Instead we will try to predict only constituent (or structure) level expression. The assumption is that this kind of expression really exists in performances and that is different and independent from note level structure. We think that a constituent level system also corresponds better to how actual human performers play music. 
%A performance redering system that only predicts note level expression would have rather meaningless fluctuations in tempo and dynamics as it does not have a notion of constituent level expression.

@@

The system will be similar to YQX in a number of ways, but with the crucial difference that expression will not be predicted per note, but per constituent. Every constituent will be played with consistend expression, the articulation, dynamics and tempo change only at constituent breaks.

@@

A structural analysis is used to derive a hierarchical structure for every score in the corpus, however, to keep the system simple we will only use this structural analysis to create a segmentation of the score into constituents. After segmentation, four score features, two of which are direct generalizations of YQX's score features, are extracted for each constituent.

So far, we have only used the score. Since we have a segmentation and every score note is associated with a performance note\footnote{In reality, not every score note is associated with a performance note since the pianist may have missed some notes. These notes will be ignored} we can also define expression per constituent. Three parameters, analogous to YQX's targets, will be used to describe expression per constituent.

@@

The resulting data is used to train a first order hidden Markov model. The system uses this model to generate performances given a score. To do this, the score is segmented into constituents, score features are extracted for each constituent. Finally viterbi decoding is used to find the sequence of expression parameters that best explains the observed score features.

The succes of a SBPR system depends largely on two factors. The ability to generate musically meaningful parse trees of a piece and the ability to accurately characterize the individual constituents and their relations with other constituents in score features. The following section address these issues.

%The following section will introduce an application of the delta framework \ref{markwin} that intends to segment a score into musically meaningfull constituents. Section \ref{scorefeatures} will deal with the characterization of individual constituents, section \ref{expressionfeatures} will explain how expressive parameters of a constituent are determined and translated into expressioin. Section \ref{performancerendering} will explain how performances are

@@

This section will describe individual components of the system sketched in section \ref{sec:approach} in more detail. The structural analysis is based on the delta framework, which will be described in section \ref{sec:deltaframework}.  Section \ref{sec:segmentation} will discuss how the delta framework is applied to get a segmentation. Sections \ref{sec:scorefeatures} and \ref{sec:targets} will describe how the scorefeatures and expression parameters are calculated. Section \ref{sec:model} will describe how a hidden Markov model is trained on our data.

@@

In his Phd thesis \cite{markwin}, Markwin van der Berg introduces a formal way of parsing music into parsetrees: the delta framework. He relates his work to the work of Lerdahl and Jackendoff \cite{lerdahl1996generative} but claims to have found a more general approach. Below, I will shortly describe the delta framework as proposed by Van der Berg.

The delta framework is based on the intuition that differences between features of notes indicate splits between constituents. The higher the difference, the higher the level of split in the parse tree (where the root note is at the highest level). Van der Berg proposes a delta rule that converts a set of differences, or deltas, between notes into a parsetree following this intuition.

@@

\[M_{ij} = [n_i, n_{i+1}, \cdots, n_j]\]

@@

The basic note features can be used to define delta functions, for example $\Delta\textbf{Pitch} = \textbf{Pitch}(n_i) - \textbf{Pitch}(n_{i-1})$. In general, a delta function $\delta(i)$ is defined as the difference of two notes in some feature $\phi$: $\delta(i) = \phi(n_i) - \phi(n_{i-1})$. We can apply a delta function to every pair of succeeding notes in a sequence to get a list of deltas: 
\[\Delta M_{ij} = [\delta(n_{i+1}), \delta(n_{i+2}), \cdots, \delta(n_{j})]\]

A recursive rule, called the delta rule can be used to translate a list of deltas into a grouping structure. This ruled, called the delta rule is shown in algorithm \ref{alg:deltarule}, where DeltaRule($M_{ij}$) is a recursive call to the algorithm itself, $M_{ij}$ is the ordered list of notes to be analyzed and $A$ is the resulting analysis.\footnote{The version here is a slightly reformulated, although functionally equivalent, version of Van den Berg's deltarule. See \cite{markwin} for his original version.}:
\begin{algorithm}
\begin{algorithmic}
\STATE{$D_{i+1,j} \leftarrow \Delta M_{ij}$}
\STATE{$A \leftarrow []$}
\STATE $m \leftarrow \textbf{ max}(D)$
\FOR{$\delta \textbf{ in } D$}
\IF{$\delta = m$}
\STATE $p = \textbf{ index of }\delta \textbf{ in } D$
\STATE $q = \textbf{ index of the next occurence of m in }D \textbf{ or } j$
\STATE $\textbf{append DeltaRule(}M_{pq}\textbf{) to } A$
\ENDIF
\ENDFOR
\RETURN{$A$}
\end{algorithmic}
\caption{The delta rule}
\label{alg:deltarule}
\end{algorithm}

The delta rule converts a sequence of notes into a nested list structure that can be interpreted as a tree. The deltarule can `parse' a piece of music into a \textit{parse tree}.

It is also possible to define higher order delta function (deltas of deltas). Second order deltafunction can be used to differentiate a group of ascending notes from a group of notes that have wildly varying pitches. Van den Berg notes that this generates ambigous grouping structures as for example a second order deltarule needs at least three notes to be specified. For three notes, three different grouping structures are possible: $[[n_1, n_2], n_3]$, $[n_1, n_2, n_3]]$ and $[n_1, [n_2, n_3]]$. Because we want to use a second order deltafunction for segmentation only, we choose to use only one interpretation: $[[n_1, n_2], n_3]$. This choice is motivated by the intuition that if in a group of three notes the delta function in some feature between the first two notes is low and the delta function between the second and third note is high (thus resulting in a high second order deltafunction) we would want a constituent split to happen between the second and third note. Note again that segmentation, like structure, may often be ambigous and multiple interpretations may be right; for our purposes it is important that at least one of these right interpretations is chosen.

Having defined first and second order deltafunctions and the deltarule, we can use first and second order deltafunctions to parse a piece of music into a parse tree. Since we have a set of six basic note features, we have a set of six interpretations(parse trees) available for a piece of music. Intuition tells us that if a particular group of notes is found in multiple parse trees, this group of notes may be eligible to be a constituent. Van den Berg captures this intuition in the form of a \textit{yield rule}. A node in the parse tree is said to `yield' a group of notes if the node recursively contains this group of notes. If two or more nodes in different trees yield the same group of notes they should be connected according to the yield rule.

Unfortunately, the yield rule addresses the problem of how to interpret multiple parse trees quite poorly. We would like to have some way of combining parsetrees into one `consensus tree'. This is not what the yield rule does. The yield rule generates a set of trees in which some nodes may be interconnected but there does not seem to be a logical way to interpret this set of interpretations and connected nodes. Van den Berg does not further address this issue. In the next section we circumvent the problem completely by not using more than one interpretation at a time.

@@

A recursive structure like the parse trees generated by the delta framework is the kind of structure that we would ideally want to attribute to music. However, because of the problems with interpreting different parsetrees and because deltatrees do not necessarily represent the kind of structure that human musical intuition would attribute to music, we chose not to use a hierarchical structure and instead settle for a segmentation, which can be more reliably derrived from deltatrees.

The goal is to find a `safe' method to segment music, that is, a method that generates segmentations that do not clash with human musical intuition. For this purpose the parse trees of delta onset (inter-onset interval, or simply \textsc{IOI}) and delta pitch (pitch interval) seem most suited: sudden jumps in pitch or onset often correspond to phrase transitions. 
\begin{figure}
\centering
\subfigure[Delta onset tree]{
\includegraphics[scale=0.4]{img/onset_sym}
\label{fig:donset}
}
\subfigure[Second order delta pitch tree]{
\includegraphics[scale=0.3]{img/second_order_pitch_sym}
\label{fig:dpitch}
}
\caption{Parse trees of the first eight bars of piano sonata KV331 III. (Turkisch March)by Mozart }
\label{fig:pitchonsettrees}
\end{figure}

A delta onset tree and a second order delta pitch tree of the first eight bars of Mozart's Turkish March can be found in figure \ref{fig:pitchonsettrees}. In his figure we can see that some nodes contain only subnodes, some nodes contain only notes and some nodes contain notes and subnodes. To translate the tree into a segmentation we use algorithm \ref{alg:segmentation} starting with the root node.
\begin{algorithm}
\caption{Segmentation}
\label{alg:segmentation}
For some node $n$, do:
\begin{enumerate}
\item If $n$ is a note, add it as a singleton constituent to the segmentation
\item Expand $n$.
\item If $n$ contains only subnodes, or less than $x$ notes, start from step 1 for every subnode and -note.
\item If $n$ contains at least $x$ notes, add the notes recursively contained by the current note to the segmentation.
\end{enumerate}
\end{algorithm}

The value of $x$ represents the tolerance of singleton constituents. We need this parameter because sometimes very long notes cause top level splits in the onset tree. We have found that setting it to the number of notes recursively contained (yielded) by the current node divided by 16 works quite well for works by Mozart. Our segmentation of the first eight bars of Mozart's Turkish March using a delta onset tree can be found in figure \ref{fig:mozartsegments}.\footnote{Note that if the delta tree from figure  \ref{fig:donset} is used the segmentation will be different. This segmentation was based on a onset delta tree over the entire piece instead of just the first eight bars}

@@

For some music like Mozart's, using only a delta onset tree works very well. However, some music, like Bach's Inventionen uses almost no differentiation in IOI. Hence, a segmentation based solely on inter onset interval will not work well for these pieces.

To overcome this problem we will use pitch. While Bach's Inventionen use almost no differentiation in IOI they can be segmented nonetheless. We will not make any attempt to combine pitch interval trees and IOI trees. Instead we take the segmentation given by the IOI tree and try to subdivide every segment using a pitch interval tree.

The sort of changes in pitch that indicate constituent transitions are not characterized very well by a first order pitch interval tree. For example a melody that has four notes that are chromatically ascending followed by four notes that have octave intervals between them should be segmented into two constituents where one contains the four ascending notes and another one contains the notes with octave intervals. A first order pitch interval tree puts the four ascending notes in one segment but puts the four octave interval notes in four separate segments. A second order pitch interval puts the split in this case exactly where we would want it.

@@

We can now convert a piece of music into a series of constituents. These constituents will be used to predict expression so we must be able to charaterize them in a way that correlates with the way they are performed. Analogous to YQX we are looking for the \textit{context} of the constituent as well as some description of the constituent itself.

@@

We can generalize pitch interval and duration ratio per note to \textit{context features}: \textbf{mean pitch interval} and \textbf{mean duration ratio}. Definitions can be found below. 
\begin{description}
\item[\textbf{Mean pitch interval}] 
The difference between the mean pitch of the current constituent and the mean pitch of the next constituent, zero if the current is the last constituent
\item[\textbf{Mean duration ratio}]
The logarithmic ratio between the mean note duration of the current constituent and the mean note duration of the next constituent
\end{description}

Since I-R arch is related to note-level expression it does not generalize well to a constituent level feature.

>>>>

% Multi line tables
\usepackage{array}
\usepackage{booktabs}

@@

% [DO NOT CHANGE] The address of the educational programme.
University of Amsterdam\\
Faculty of Science\\
Science Park 904\\
1098 XH Amsterdam

@@

% [CHANGE] The address of the institute at which your supervisor is working.
% Be sure to include (1) institute (is appropriate), (2) faculty (if
% appropriate), (3) organization name, (4) organization address (2 lines).
Institute for Language and Logic\\
Faculty of Science\\
University of Amsterdam\\
Science Park 904\\
1098 XH  Amsterdam

@@

%Musical structure is extracted automatically. A dataset containing performances aligned to scores is then used to learn how structure relates to expression. A hidden Markov model is used to model this relation. Finally, the system can be applied to a a new score to produce an expressive performance.

@@

{\bf Keywords:} performance rendering, musical structure, constituent structure
\end{abstract}
\section{Introduction}

@@

Given a hypothetical perfect synthesizer, performing scores with computers is a trivial task. However, \textit{expressively} performing music is not, and much research has focussed on this issue. A wide selection of scientific fields may have interest in such research: Artificial intelligence, cognitive science and musicology all benefit from a better understanding of expression in music.

@@

A rather practical approach to expressive music performance is to recreate it. A performance rendering system is a computer program that automatically creates an expressive, human-like performance of a musical score. Performance rendering systems have always faced a problem of evaluation. It is debatable how much we should value correlation of generated performances with human performances of the same piece. A Chopin interpretation by Arthur Rubinstein can be radically different from a Chopin interpretation by Vladimir Horowitz, yet both are considered high quality interpretations. This problem has been one of the motivations for a yearly performance rendering event where performance rendering systems compete and are judged by an audience \cite{hiraga2002rencon}.

Over the years, many performance rendering systems have been proposed. Early attempts, like Director Musices \cite{friberg2000generating}, have applied sets of rules to the problem. As the availability of corpora grew, machine learning systems have been proposed as well. Gerhard Widmer is working with a large dataset containing almost all works by Chopin, performed by one pianist. A performance rendering system using this dataset called YQX \cite{yqx} was proposed by Sebastian Flossman et al. in 2008. This system won in al three categories of Rencon2008. At Rencon2011 it was ranked best in a blind evaluation by experts and ranked second in the evaluation by an audience. It should be noted that despite its age, Director Musices ranked third in the blind evaluation, suggesting that although performance rendering systems have advanced, the field is still in its infancy.

Many of the performance rendering systems that use machine learning, including YQX, predict expression at note level. Some do use structure. Pop-E \cite{hashida2006pop}, a rule based performance rendering system, uses a structure analysis based on Lerdahl and Jackendoff's grouping preference rules \cite{lerdahl1996generative}, but the structural analysis must be supplied to it by a human expert. In this thesis we will motivate the need for structure level expression and propose a system that purely uses structure level expression. The YQX system is very successful at the moment, therefore we will use this system as a reference for introducing our own system. In the following section we will provide a short analysis of some shortcomings of the YQX system and argue what we believe is the reason for these shortcomings.

What we mean by structure level expression and why we think it is important is clarified in section \ref{sec:structure}. Section \ref{sec:approach} will describe the system as a whole. Section \ref{sec:method} will describe every component of the system in detail. Practical issues are discussed in section \ref{sec:implementation}. Results are presented in section \ref{sec:results} and evaluated in section \ref{sec:discussion}. Suggestions for improving the system and future research can be found in section \ref{sec:futurework}.

%Early performance rendering systems used rule-based approaches to recognize small local structures that are used as indications to add certain kind of expression. A prominent ongoing rule based system, \textit{Director Musices} uses 30 rules with weighting parameters that take music features as input and produce expressive actions as output. The rules where created with the help of musical experts. \textit{Pop-E} is a more recent rule based system that uses aspects of Lerdahl and Jackendoff's Generative Theory of Tonal Music(GTTM) [Introduce this]. Gerhard Widmer is doing research using a large dataset of performances aligned to scores.

@@

The YQX system, as Flossman et al. admit, tended to sometimes produce nervous sounding changes in expression. They present two extensions tailored towards generating smoother performances. The problem with these extensions is, as Flossman et al. also admit themselves, that the increased smoothness comes at the expense of expressivity. To compensate for this, three explicit rules are added to post-process the performances. We think the reason that Flossman et al. stumbled upon this tradeoff between expressiveness and nervousness is that the nervousness is inherent to note level performance rendering. Note level systems simply are not suited to capture one component of expression, namely structure level expression.

The system presented here will use structure exclusively to generate performances and will completely ignore note level expression. We hope that this system is able capture the sudden changes of expression that a note level system cannot easily learn. Sudden changes in expression occur frequently in expressive performances, a few notes may be played very loud, or very slow followed by soft notes played very fast. When done at the right moment, such sudden changes can greatly improve the human-like feel of the performance, however when done at the wrong moment, they can ruin a performance. We think that at structure level it will be easier to capture the context in which sudden changes occur.

@@

\subsection{Structure Level Expression}
When listening to music, the listener's musical intuition assigns a certain hierarchical structure to the music: Notes make up phrases, phrases make up themes and themes make up a piece. In a performance, this structure may be accentuated in different ways. Accentuation happens at different levels, at note level performers may slow down at the end of a phrase or introduce small pauses in the performance at phrase transitions. At constituent level one phrase may be played very loud, fast or staccato, while the next may be played slow, soft and legato.

To formally describe musical structure, we can look at music in a way similar to the way we look at natural language processing(NLP). In this analogy we see a piece of music as a sentence, which consists of constituents that individually can be made of constituents as well. We can recursively subdivide constituents into more constituents until we reach a terminal symbol. In NLP this may be a word, in music, this may be a note. We could represent musical structure as a parse tree. This paradigm corresponds to the intuition that a melody is not simply a sequence of notes but that notes form phrases. A phrase is always a constituent but a constituent is not always a phrase, it can be a distinguished part of a phrase or a distinguished set of phrases as well.

We must note that musical scores can be highly ambiguous and even experienced listeners may not agree on the correct parse tree of a piece. Quite often there may simply be more than one parse tree that makes musical sense. This should not be a problem for a performance rendering system: different expressive interpretations of one piece can be very diverse and still be accepted as sensible interpretations of the piece. As long as the parse tree does make at least some musical sense, a performance rendering system should be able to use it.

Although the YQX does have some notion of structure\footnote{One of YQX's note features is distance to nearest point of closure, see section \ref{sec:scorefeatures}.}, expression is only predicted per note. The authors admit that the first simple version of the system ``tended to sometimes produce unstable, `nervous' sounding performances''. The only way to overcome this problem was to introduce methods that limited the expressiveness of performances. We consider this trade-off to be inherent to note-level expression based systems. To solve it, some notion of structure level expression is required.

@@

In this thesis, we propose a structure based performance rendering (SBPR) system. The system presented here ignores note level expression. Instead we will try to predict only constituent (or structure) level expression. The assumption is that this kind of expression really exists in performances and that it is different and independent from note level structure. We think that a constituent level system also corresponds better to how actual human performers play music. 
%A performance rendering system that only predicts note level expression would have rather meaningless fluctuations in tempo and dynamics as it does not have a notion of constituent level expression.

@@

The system will be similar to YQX in a number of ways, but with the crucial difference that expression will not be predicted per note, but per constituent. Every constituent will be played with consistent expression, the articulation, dynamics and tempo change only at constituent breaks.

@@

A structural analysis is used to derive a hierarchical structure for every score in the corpus; however, to keep the system simple we will only use this structural analysis to create a segmentation of the score into constituents. After segmentation, four score features, two of which are direct generalizations of YQX's score features, are extracted for each constituent.

So far, we have only used the score. Since we have a segmentation and every score note is associated with a performance note\footnote{In reality, not every score note is associated with a performance note since the pianist may have missed some notes. These notes will be ignored.} we can also define expression per constituent. Three parameters, analogous to YQX's targets, will be used to describe expression per constituent.

@@

The resulting data is used to train a first-order hidden Markov model. The system uses this model to generate performances given a score. To do this, the score is segmented into constituents, score features are extracted for each constituent. Finally Viterbi decoding is used to find the sequence of expression parameters that best explains the observed score features.

The success of a SBPR system depends largely on two factors. The ability to generate musically meaningful parse trees of a piece and the ability to accurately characterize the individual constituents and their relations with other constituents in score features. The following section addresses these issues.

%The following section will introduce an application of the delta framework \ref{markwin} that intends to segment a score into musically meaningful constituents. Section \ref{scorefeatures} will deal with the characterization of individual constituents, section \ref{expressionfeatures} will explain how expressive parameters of a constituent are determined and translated into expression. Section \ref{performancerendering} will explain how performances are

@@

This section will describe individual components of the system sketched in section \ref{sec:approach} in more detail. The structural analysis is based on the delta framework, which will be described in section \ref{sec:deltaframework}.  Section \ref{sec:segmentation} will discuss how the delta framework is used to get a segmentation. Sections \ref{sec:scorefeatures} and \ref{sec:targets} will describe how the score features and expression parameters are calculated. Section \ref{sec:model} will describe how a hidden Markov model is trained on our data.

@@

In his Ph. D. thesis \cite{markwin}, Markwin van den Berg introduces a formal way of parsing music into parse trees: the delta framework. He relates his work to the work of Lerdahl and Jackendoff's Generative Theory of Tonal Music \cite{lerdahl1996generative} but claims to have found a more general approach. Below, I will shortly describe the delta framework as proposed by Van den Berg.

The delta framework is based on the intuition that differences between features of notes indicate splits between constituents. The larger the difference, the higher the level of the split in the parse tree (where the root note is at the highest level). Van den Berg proposes a delta rule that converts a set of differences, or deltas, between notes into a parse tree following this intuition.

@@

\[M_{i,j} = [n_i, n_{i+1}, \cdots, n_j]\]

@@

The basic note features can be used to define delta functions, for example $\delta_{\textbf{Pitch}} = \textbf{Pitch}(n_i) - \textbf{Pitch}(n_{i-1})$. In general, a delta function $\delta(i)$ is defined as the difference of two notes in some feature $\phi$: $\delta_{\phi}(i) = \phi(n_i) - \phi(n_{i-1})$. We can apply a delta function to every pair of succeeding notes in a sequence to get a list of deltas: 
\[\Delta_{\phi} M_{i,j} = [\delta_{\phi}(n_{i+1}), \delta_{\phi}(n_{i+2}), \cdots, \delta_{\phi}(n_{j})]\]

A recursive rule, called the delta rule, can be used to translate a list of deltas into a grouping structure. This rule, called the delta rule is shown in algorithm \ref{alg:deltarule}, where DeltaRule($\phi, M_{i,j}$) is a recursive call to the algorithm itself, $\phi$ is the feature that we use for the delta function, $M_{i,j}$ is the ordered list of notes to be analyzed and $A$ is the resulting analysis.\footnote{The version here is a slightly reformulated, although functionally equivalent, version of Van den Berg's delta rule. See \cite{markwin} for his original version.} The algorithm is initialized with the entire piece of music, $M_{0, n}$ and the feature, $\phi$, to use for the delta function: DeltaRule($\phi, M_{0,n}$).
\begin{algorithm}
\begin{algorithmic}
\STATE{$D_{i+1,j} \leftarrow \Delta_{\phi} M_{i,j}$}
\STATE{$A \leftarrow []$}
\STATE $m \leftarrow \mbox{ max}(D)$
\STATE $p \leftarrow i$
\FOR{$d \mbox{ in } D$}
\IF{$d = m$}
\STATE $q \leftarrow \mbox{ index of }d \mbox{ in } D$
\STATE $\mbox{\textbf{append} DeltaRule(}\phi, M_{p,q}\mbox{) to } A$
\STATE $p \leftarrow q$
\ENDIF
\ENDFOR
\STATE $\mbox{\textbf{append} DeltaRule(}\phi, M_{p,j}\mbox{) to } A$
\RETURN{$A$}
\end{algorithmic}
\caption{The delta rule}
\label{alg:deltarule}
\end{algorithm}

The delta rule converts a sequence of notes into a nested list structure that can be interpreted as a tree. The delta rule parses a piece of music into a \textit{parse tree}. The resulting parse tree depends on what feature $\phi$ we used for the delta function. Every feature yields a different parse tree.

It is also possible to define higher order delta function (deltas of deltas). Such higher order functions can detect patterns that would remain undetected by first-order delta function. Second order pitch delta functions, for example, can be used find constituent splits between a group of steadily ascending notes and a group of notes that have wildly varying pitches.

%Van den Berg notes that this generates ambiguous grouping structures as for example a second order delta rule needs at least three notes to be specified. For three notes, three different grouping structures are possible: $[[n_1, n_2], n_3]$, $[n_1, n_2, n_3]]$ and $[n_1, [n_2, n_3]]$. Because we want to use a second order delta function for segmentation only, we choose to use only one interpretation: $[[n_1, n_2], n_3]$. This choice is motivated by the intuition that if in a group of three notes the delta function in some feature between the first two notes is low and the delta function between the second and third note is high (thus resulting in a high second order delta function) we would want a constituent split to happen between the second and third note. Note again that segmentation, like structure, may often be ambiguous and multiple interpretations may be right; for our purposes it is important that at least one of these right interpretations is chosen.

%Having defined first and second order delta functions and the delta rule, we can use first and second order delta functions to parse a piece of music into a parse tree.

Since we have six basic note features, we have six first-order delta functions and six interpretations (parse trees) for a piece of music. Although each parse tree provides us musically relevant information, we would like to have only one tree. Intuition tells us that if a particular group of notes is found in multiple parse trees, this group of notes may be eligible to be a constituent. Van den Berg captures this intuition in the form of a \textit{yield rule}. A node in the parse tree is said to `yield' a group of notes if the node recursively contains this group of notes. If two or more nodes in different trees yield the same group of notes they should be connected according to the yield rule.

Unfortunately, the yield rule addresses the problem of how to interpret multiple parse trees quite poorly. We would like to have some way of combining parse trees into one `consensus tree'. This is not what the yield rule does. The yield rule generates a set of trees in which some nodes may be interconnected but there does not seem to be a logical way to interpret this set of interpretations and connected nodes. Van den Berg does not further address this issue. In the next section we circumvent the problem completely by not using more than one interpretation at a time.

@@

A recursive structure like the parse trees generated by the delta framework is the kind of structure that we would ideally want to attribute to music. However, because of the problems with interpreting different parse trees and because delta trees do not necessarily represent the kind of structure that human musical intuition would attribute to music, we chose not to use a hierarchical structure and instead settle for a segmentation, which can be more reliably derived from delta trees.

The goal is to find a `safe' method to segment music, that is, a method that generates segmentations that do not clash with human musical intuition. For this purpose the parse trees of delta onset (inter-onset interval, or simply \textsc{IOI}) and delta pitch (pitch interval) seem most suited: sudden jumps in pitch or onset often correspond to phrase transitions. 
\begin{figure}
\centering

\subfigure[Delta onset tree]{
\includegraphics[scale=0.4]{img/onset_sym}
\label{fig:subfig1}
}

\subfigure[Second order delta pitch tree]{
\includegraphics[scale=0.3]{img/second_order_pitch_sym}
\label{fig:subfig2}
}
\caption{Parse trees of the first eight bars of piano sonata KV331 III. (Turkisch March)by Mozart }
\label{fig:pitchonsettrees}
\end{figure}

A delta onset tree and a second order delta pitch tree of the first eight bars of Mozart's Turkish March can be found in figure \ref{fig:pitchonsettrees}. In this figure we can see that some nodes contain only other nodes, some nodes contain only notes and some nodes are mixed and contain notes and other nodes. We will use this differentiation between node types in the translation of a tree into a segmentation, which is shown in algorithm \ref{alg:segmentation}. The algorithm is initialized with the root node.
\begin{algorithm}
\caption{Segmentation}
\label{alg:segmentation}
For some node $n$, do:
\begin{enumerate}
\item \textbf{If} $n$ is a note.\\
\textbf{Then} add $n$ as a singleton segment to the segmentation.\\
 \textbf{Else} expand $n$.
\item \textbf{If} $n$ is mixed and directly contains more than $x$ notes.\\
\textbf{Then} add all notes recursively contained by $n$ as a segment to the segmentation.\\
\textbf{Else} start from step 1 for every child of $n$.
% If $n$ contains at least $x$ notes, add the notes recursively contained by $n$ to the segmentation.
\end{enumerate}
\end{algorithm}

The value of $x$ represents the tolerance of singleton segments. We need this parameter because, in the delta onset tree, very long notes cause one note to split of a constituent at a high level, in this case we want to accept that this one note will be a singleton segment. We have found that setting $x$ to a variable number, namely: the number of notes recursively contained (yielded) by the current node divided by 16 works quite well for works by Mozart. Our segmentation of the first eight bars of Mozart's Turkish March using a delta onset tree can be found in figure \ref{fig:mozartsegments}.\footnote{Note that if the delta tree from figure  \ref{fig:subfig2} is used the segmentation will be different. This segmentation was based on an onset delta tree over the entire piece instead of just the first eight bars.}

@@

For some works, like most works by Mozart, using only a delta onset tree works very well. However, for other, like Bach's Inventionen, a delta onset tree is not sufficient since the music contains almost no differentiation in IOI (most notes are equidistant in onset).

To overcome this problem we will use pitch. Even though Bach's Inventionen use almost no differentiation in IOI they can be segmented nonetheless. We will not make any attempt to combine pitch interval trees and IOI trees. Instead we take the segmentation given by the IOI tree and try to subdivide every segment using a pitch interval tree.

The sort of changes in pitch that indicate constituent transitions are not characterized very well by a first-order pitch interval tree. Let us take the example from the previous section, where we have a melody of eight notes, four of which are chromatically ascending followed by four notes that have octave intervals between. We want this to be segmented into two constituents where one contains the four ascending notes and another one contains the notes with octave intervals. A first-order pitch interval tree puts the four ascending notes in one segment but puts the four octave interval notes in four separate segments. A second order pitch interval puts the split in this case exactly where we would want it.

@@

We can now convert a piece of music into a series of constituents. These constituents will be used to predict expression, so we must be able to characterize them in a way that correlates with the way they are performed. Analogous to YQX we are looking for the \textit{context} of the constituent, but in contrast to YQX we also need some description of the constituent itself.

@@

We can generalize pitch interval and duration ratio per note to \textit{context features}: \textbf{mean pitch interval} and \textbf{mean duration ratio}. Definitions can be found below. 
\begin{description}
\item[\textbf{Mean pitch interval}] 
The difference between the average pitch of the current constituent and the average pitch of the next constituent, zero if the current is the last constituent
\item[\textbf{Mean duration ratio}]
The logarithmic ratio between the average note duration of the current constituent and the average note duration of the next constituent
\end{description}

Since I-R arch is related to note-level expression, it does not generalize well to a constituent level feature.
